---
title: "loadNprocess"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{loadNprocessR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup}
# devtools::github_install("kvitense/LM_Chla")
library(LMChla)
```

Note: This requires the package "devtools" which can be installed as `r install.packages("devtools")`.

# Acquire data
```{r acquireData}
data("lakeMichigan") # this will result in object called "lakeMichigan" in your global environment
```
Note: That this comes from a frozen realization of the source data and therefore may not be up to date.


OR Assembling using functions provided by the package
```{r}
lakeMichigan <- assembleData("filepath")
```
- This will save an Rda binary version of the compiled data to the location specified by "filepath"
  - Rds files are loadable using `df <- load("filepath")` in R.

# Suggested workflows
Suggested workflows (still in development) are included with the package. [These workflows](R/postProcessing.R) provide the ability to:

- Impute censored data based on detection limits
  - starting with data (`df`) in long format
```{r}
lakeMichigan <- .dlImputation(lakeMichigan, imputeMethod = "halfMDL")
```

- Pivot from long to wide format
```r
lakeMichigan <- .exactPivot(lakeMichigan)
```
- Impute missing observations based on a nearest neighbor search
- Naively impute remaining missingness using a random forest model