---
title: "Data Checks"
date: today
---

```{r setup}
library(tidyverse)
library(kableExtra)
df <- readRDS("Data/fulldata.rds")
```


# Missingness
```{r}
df %>%
  select(1:9, 16:21, 23) %>%
  reframe(across(everything(), function(x) mean(is.na(x))), .by = Study) %>%
  kable() %>%
  kableExtra::kable_paper()
```

Notes on missingness:
- GLENDA
  - Has ~50% missingness
- SeaBird
  - Can't locate station depth in their files
- NCCA
  - Secchi missing sample depth because it is an integrated sample (over water column)

- CSMI
  - 2010
    - Lat/Lon reported only to nearest whole number 
      - there are real Lat/ real Lon in smpstts10.xls, but not sure what they mean
    - Lat/Lon/depth/sampleDate partially missing because odd reporting in file
      - Can we use forward filling to fill in the missing values?
    - Can't find sampleDepth
      - Just report the water layer

  - 2015 
    - It actually just has a lot of missing WQ 
      - I'm guessing this is because if CTD was measured and WQ wasn't they still record it as a NA
  - 2021
    - actually has ~50% missingness


```{r}
GLENDA %>%
  filter(Study == "GLENDA", is.na(RESULT)) %>%
  count(CodeName, ConversionFactor, is.na(as.numeric(RESULTstart))) %>%
  print(n = 64)
```
